Optimal Light Curve Attitude Inversion with Measurement Noise: Two Case Studies

### 2025-03-12 paper.tex
fixed most comments for this round

If no initial attitude state is available, as in most cases of passive debris observation, the inversion problem must be solved globally. Filtering approaches can be applied here too; particle \cite{linares2014particle, holzinger2014} and Bayesian multi-hypothesis filters \cite{burton2021two, cabrera2023} have shown promise in prior work. If a reliable prior distribution for the attitude state is not available -- which it generally is not -- a Bayesian method may be inadvisable. In the absence of prior knowledge about the object's attitude state, simulated annealing \cite{gagnon2024, clark2020}, genetic algorithms \cite{gagnon2024, piergentili2017, clark2020}, and particle swarm optimizers (PSOs) \cite{clark2020, clark2022, burton2024journal, burton2024scitech, gagnon2024} can be applied to search the solution space exhaustively.

Much of the past literature on attitude inversion from real light curve data is centered on determining the constant spin rate and axis of rotation of the observed object. Frequency analysis in the form of periodograms, Fourier transforms, or epoch folding, is often used to determine the spin rate based on the light curve's frequency spectrum \cite{silha2015, silha2021, isoletta2024, schildknecht2015, pittet2017, yanagisawa2012, koshkin2018}. If optical material properties are known, the width of a single specular glint can provide sufficient information to determine the rotation rate \cite{hinks2016}. There has also been interest in tracking the evolution of spin rates over time for different classes of objects in different orbital regimes \cite{koshkin2018, blacketer2022, karpov2016}. These spin rate determination methods are designed for single-axis rotations and can fail for tumbling objects.

@@

Our method is a global full-state attitude solver inspired by recent PSO work \cite{burton2024journal} but is easier to tune, can be fully parallelized, and is robust to significant measurement noise \cite{robinson2025att}. Further, our method differs from other global estimation methods in that we do not seek to find one optimal solution and instead search for a collection of high-quality local minima to directly study the many ambiguous solutions for a given light curve. To solve the optimization problem, we sample many candidate solutions scattered throughout the solution space and perform gradient-based optimization on each using the BFGS algorithm. Our optimization procedure reliably converges to high-quality maximum likelihood estimates of the attitude state. The final output of our method is an array of the best solutions, ranked by their likelihood. Clusters of low-error minima can be analyzed to identify families of solutions.

This work proceeds by describing the inversion method, discussing efficient methods for computing the time-varying probability density of the light curve, and presenting inversion results for the two test cases.

@@

@@

\begin{equation} \label{eq:nll_loss}
 f(\vctr{x}) = \frac{1}{m}\sum_{k=1}^{m}\left[\ln\sigma_k + \frac{1}{2}\left(\frac{S_k - \hat{S}_k(\vctr{x})}{\sigma_k}\right)^2 \right].
 \end{equation}

@@

The state vector $\vctr{x}$ we optimize to fit the observed light curve is the concatenation of the Modified Rodriguez Parameter (MRP) $\vctr{p} = [p_1, p_2, p_3]^T$, the principal body frame angular velocity $\vctr{\omega} = [\omega_1, \omega_2, \omega_3]$ in radians per second, and the ratios of the principal moments of inertia $J_y / J_x$ and $J_z / J_x$ where the inertia tensor in principal axes is $J = \mathrm{diag}\left([J_x, J_y, J_z]\right)$:

@@

The bidirectional reflectance distribution function (BRDF) defines the amount of incident radiation from $\unitv{s}$ reflected per steradian in the observer's direction $\unitv{o}$. We choose the Blinn-Phong \cite{blinn1977} BRDF for this work as it is efficient to compute and satisfies the three main requirements for a physically meaningful BRDF as it is nonnegative, energy-conserving, and reciprocal \cite{duvenhage2013}. We avoid microfacet models for computational efficiency, and because in practice, the uncertainty in the reflective properties of the observed objects will often be larger than errors introduced by the choice of reflection model. The Blinn-Phong BRDF is parameterized by the coefficient of diffuse reflectivity $C_d$, the coefficient of specular reflectivity $C_s$, and the specular exponent $n>0$:

@@

@@

To demonstrate the effectiveness of our inversion method, we first present results for a simulated light curve where the ground truth is known and material property mismatch is insignificant. We produce a noisy, observed light curve for a high-fidelity model of the Saturn V second stage scaled down by a factor $10/3$ to act as a template rocket body object, shown in Figures \ref{fig:meas_model}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\figsmall]{rocket_bodies2.png}
  \caption{Measured rocket body model \cite{nasa_models}.}
  \label{fig:meas_model}
\end{figure}

@@

Simulating the mean light curve and its approximate probability distribution via Equation \ref{eq:lc_dist} produces the light curve displayed in ADU in Figure \ref{fig:obs_adu_synth} and V-band magnitude in Figure \ref{fig:obs_mag_synth}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\figmed]{light_curves_adu_rocket_body_sdc.png}
  \caption{Simulated observed flux for the rocket body.}
  \label{fig:obs_adu_synth}
\end{figure}

@@

To emulate the reality that the precise geometry of the observed object is usually uncertain, we use perform attitude inversion with a significantly simplified rocket body model, shown in Figure \ref{fig:inv_model}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\figmed]{rocket_bodies1.png}
  \caption{Simplified rocket body model used for inversion.}
  \label{fig:inv_model}
\end{figure}

A total of $10^5$ initial state guesses were created via the sampling scheme described in Section \ref{sec:run_solver} with $\norm{\vctr{\omega}}_\text{max}=5.26$ degrees per second. The inertia tensor ratios were initialized at the truth, but are allowed to be changed during optimization. BFGS minimization was then performed for each initial guess to locally solve the optimization problem given by Equation \ref{eq:opt_problem}. Classifying candidate solutions with $\ell = 1/10$ produces over $10^4$ solutions within this bound. Figure \ref{fig:conv_lcs_synth} shows the mean light curves produced by the identified candidate solutions. There is generally good agreement between the candidate light curves and the observations, indicating that high-quality local minima have been identified.

>>>>

If no initial attitude state is available, as in most cases of passive debris observation, the inversion problem must be solved globally. Filtering approaches can be applied here too; particle \cite{linares2014particle, holzinger2014} and Bayesian Multi-Hypothesis filters \cite{burton2021two, cabrera2023} have shown promise in prior work. Bayesian methods may be ineffective if a reliable prior distribution for the attitude state is not available. In the absence of prior knowledge about the object's attitude state, simulated annealing \cite{gagnon2024, clark2020}, genetic algorithms \cite{gagnon2024, piergentili2017, clark2020}, and particle swarm optimizers (PSOs) \cite{clark2020, clark2022, burton2024journal, burton2024scitech, gagnon2024} can be applied to search the solution space exhaustively.

While much work has performed full state attitude inversion on simulated light curve data \cite{burton2024journal, gagnon2024, robinson2025att}, studies using real data often focus on determining a constant spin rate and axis of rotation for the observed object. Frequency analysis in the form of periodograms, Fourier transforms, or epoch folding, is often used to determine the spin rate based on the light curve's frequency spectrum \cite{silha2015, silha2021, isoletta2024, schildknecht2015, pittet2017, yanagisawa2012, koshkin2018}. If optical material properties are known, the width of a single specular glint can provide sufficient information to determine the rotation rate \cite{hinks2016}. There has also been interest in tracking the evolution of spin rates over time for different classes of objects in different orbital regimes \cite{koshkin2018, blacketer2022, karpov2016}. These spin rate determination methods are designed for single-axis rotations and can fail for tumbling objects.

@@

Our method is a global full-state attitude solver inspired by recent PSO work \cite{burton2024journal} but is easier to tune, can be fully parallelized, and is robust to significant measurement noise \cite{robinson2025att}. Further, our method differs from other global estimation methods in that we do not seek to find one optimal solution and instead search for a collection of high-quality local minima to directly study the many ambiguous solutions for a given light curve. Our approach was designed specifically for use with real data as we take time-varying measurement noise and uncertainties in the body moments of inertia into account explicitly.

To solve the optimization problem, we sample many candidate solutions scattered throughout the solution space and perform gradient-based optimization on each using the BFGS algorithm. Our optimization procedure reliably converges to high-quality maximum likelihood estimates of the attitude state. The final output of our method is an array of the best solutions, ranked by their likelihood. Clusters of low-error minima can be analyzed to identify families of solutions.

This work proceeds by describing the inversion method, discussing efficient methods for computing the time-varying probability density of the light curve. We present inversion results for a simulated rocket body light curve where the ground truth is known, followed by two cases -- a derelict satellite and rocket body -- using real light curve data collected by the Purdue Optical Ground Station. A deeper treatment of our method and additional simulated light curve results can be found in a forthcoming publication by the authors \cite{robinson2025att}.

@@

The state vector $\vctr{x}$ we optimize to fit the observed light curve is the concatenation of the Modified Rodriguez Parameter (MRP) $\vctr{p} = [p_1, p_2, p_3]^T$, the principal body frame angular velocity $\vctr{\omega} = [\omega_1, \omega_2, \omega_3]$ in radians per second, and the ratios of the principal moments of inertia $J_y / J_x$ and $J_z / J_x$ where the constant inertia tensor in principal axes is $J = \mathrm{diag}\left([J_x, J_y, J_z]\right)$:

@@

\begin{equation} \label{eq:nll_loss}
 f(\vctr{x}) = \frac{1}{m}\sum_{k=1}^{m}\left[\ln\sigma_k + \frac{1}{2}\left(\frac{\vctr{S}_k - \hat{\vctr{S}}_k(\vctr{x})}{\sigma_k}\right)^2 \right].
 \end{equation}

@@

@@

The bidirectional reflectance distribution function (BRDF) defines the amount of incident radiation from $\unitv{s}$ reflected per steradian in the observer's direction $\unitv{o}$. We choose the Blinn-Phong \cite{blinn1977} BRDF for this work as it is efficient to compute and satisfies the three main requirements for a physically meaningful BRDF as it is nonnegative, energy-conserving, and reciprocal \cite{duvenhage2013}. We avoid microfacet models for computational efficiency, and because in practice, the uncertainty in the reflective properties of the observed objects will often be larger than errors introduced by the choice of reflection model. The Blinn-Phong BRDF is parameterized by the coefficient of diffuse reflectivity $C_d$, the coefficient of specular reflectivity $C_s$, and the specular exponent $n>0$ \cite{duvenhage2013}:

@@

This section summarizes an in-depth study of relevant noise sources in CCD and CMOS sensor light curves by the authors in a forthcoming publication \cite{robinson2025twin}.

@@

To demonstrate the effectiveness of our inversion method, we first present results for a simulated light curve where the ground truth is known and material property mismatch is insignificant. We produce a noisy, observed light curve for a high-fidelity model of the Saturn V second stage scaled down by a factor $10/3$ to act as a template rocket body object, shown on the right of Figure \ref{fig:rb_models}.

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=\figsmall]{rocket_bodies2.png}
%   \caption{Simulated truth rocket body model \cite{nasa_models}.}
%   \label{fig:meas_model}
% \end{figure}

@@

Simulating the mean light curve and its approximate probability distribution via Equation \ref{eq:lc_dist} produces the light curve displayed in V-band magnitude in Figure \ref{fig:obs_mag_synth}.

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=\figmed]{light_curves_adu_rocket_body_sdc.png}
%   \caption{Simulated observed flux for the rocket body.}
%   \label{fig:obs_adu_synth}
% \end{figure}

@@

To emulate the reality that the precise geometry of the observed object is usually uncertain, we use perform attitude inversion with a significantly simplified rocket body model, shown in Figure \ref{fig:rb_models}.

\begin{figure}[H]
  \centering
  \includegraphics[width=\figmed]{rocket_bodies.png}
  \caption{Simplified rocket body model used for inversion (left) and simulated ground truth model (right).}
  \label{fig:rb_models}
\end{figure}

A total of $10^5$ initial state guesses are created via the sampling scheme described in Section \ref{sec:run_solver} with $\norm{\vctr{\omega}}_\text{max}=5.26$ degrees per second. The inertia tensor ratios are initialized by sampling a symmetric Gaussian distribution with $\sigma=0.05 \: [kg\cdot m^2]$ on each axis to simulate a moderate level of inertia uncertainty. BFGS minimization was then performed for each initial guess to locally solve the optimization problem given by Equation \ref{eq:opt_problem}. Classifying candidate solutions with $\ell = 1/10$ produces over $10^4$ solutions within this bound. Figure \ref{fig:conv_lcs_synth} shows the mean light curves produced by the identified candidate solutions. There is generally good agreement between the candidate light curves and the observations, indicating that high-quality local minima have been identified.
